Notes on Validated Model Counting

### 2021-04-05 doc/mc-notes.tex
Minor updates

\begin{lemma}
\label{lemma:independent}
if Boolean formula $F$ is independent of variable $x$, then:
\begin{equation}
|\interpset{F}| \; = \; 2\cdot |\interpset{F\land x}| = \; 2\cdot |\interpset{F\land \obar{x}}| \label{eqn:independent}
\end{equation}  
\end{lemma}  
That is, the set of interpretations of $F$ are split evenly between
those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.

@@

We consider only graphs having unique root nodes, i.e., nodes with no outgoing arcs.

@@

We can define the function $\depend$ mapping each node
$v$ to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \{ x \} \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

\begin{lemma}
\label{lemma:free}
Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
\end{lemma}  
\noindent
{\em Proof:} Follows by the recursive definition of $\depend$ (\ref{eqn:depend:true}--\ref{eqn:depend:ite}) and by (\ref{eqn:ite:interp}).

A {\em free} ITE graph is a proper ITE graph satisfying the added
restriction that for every chain of nodes from an input or constant
node to the root node, and for every variable $x \in X$, only a single
operator node can have the input node for $x$ as its \ifarg{} input.
More formally, for every operator node $v = \ite(x, v_t, v_e)$ in a
free ITE graph, its incoming \thenarg{} and \elsearg{} arcs must be
from nodes such that $x \not \in \depend(v_t)$ and $x \not \in
\depend(v_e)$.

@@

\begin{theorem}
Equations (\ref{eqn:count:true}--\ref{eqn:count:ite}) correctly hold for any node $v$ in a free ITE graph.
\end{theorem}  
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemma~\ref{lemma:independent}, (\ref{eqn:ite:interp}),
Lemma~\ref{lemma:free}, and (\ref{eqn:depend:ite}).  That is, both
$v_t$ and $v_e$ are independent of $x$.  That implies that half of the
satisfying assignments $\interp$ for $v_t$ have have $\interp(x) =
\tautology$.  Similarly, half of the satisfying assignments $\interp$
for $v_e$ have $\interp(x) = \nil$.  Equation (\ref{eqn:count:ite}) then follows by (\ref{eqn:ite:interp}).

>>>>

\begin{lemma}
\label{lemma:independent:split}
if Boolean formula $F$ is independent of variable $x$, then:
\begin{equation}
|\interpset{F}| \; = \; 2\cdot |\interpset{F\land x}| = \; 2\cdot |\interpset{F\land \obar{x}}| \label{eqn:independent}
\end{equation}  
\end{lemma}  
That is, the set of interpretations of $F$ are split evenly between
those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.

@@

@@

We can define the function $\depend$ mapping each node
$v$ in a proper ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \{ x \} \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

\begin{lemma}
\label{lemma:independent:dset}
Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
\end{lemma}  
\noindent
{\em Proof:} Follows by the recursive definition of $\depend$ (\ref{eqn:depend:true}--\ref{eqn:depend:ite}) and by (\ref{eqn:ite:interp}).

A {\em free} ITE graph is a proper ITE graph satisfying the added
property that for every operator node $v = \ite(x, v_t, v_e)$, its
incoming \thenarg{} and \elsearg{} arcs must be from nodes such that
$x \not \in \depend(v_t)$ and $x \not \in \depend(v_e)$.

Free ITE graphs correspond directly to a class of binary decision diagrams known as {\em free} BDDs (FBDDs)~\cite{gergov:ieeetc:1994}.  Ordered BDDs are a subclass of FBDDs.

@@

\begin{theorem}
Equations (\ref{eqn:count:true}--\ref{eqn:count:ite}) correctly hold for any node $v$ in a free ITE graph.
\end{theorem}  
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemma~\ref{lemma:independent:split}, (\ref{eqn:ite:interp}),
Lemma~\ref{lemma:independent:dset}, and (\ref{eqn:depend:ite}).  That is, both
$v_t$ and $v_e$ are independent of $x$.  That implies that half of the
satisfying assignments $\interp$ for $v_t$ have have $\interp(x) =
\tautology$.  Similarly, half of the satisfying assignments $\interp$
for $v_e$ have $\interp(x) = \nil$.  Equation (\ref{eqn:count:ite}) then follows by (\ref{eqn:ite:interp}).
